Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-06-28 Documents/M1/realizableFermionicM1/sections/appendix.tex
Minor edits

\newpage 
\section{Construction of IMEX Scheme PD-ARS}
\label{app:PD-ARS}

Here we construct a three-stage PD-IMEX scheme of Type~ARS, conforming to Definition~\ref{def:PD-IMEX}.  
We refer to the resulting IMEX scheme as PD-ARS.  
The double Butcher tableau is
\begin{equation}
  \begin{array}{c | c c c}
  	         0           & 0                 & 0                   & 0  \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0                   & 0  \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0  \\ \hline
  	                   & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0  & 0  & 0         & 0          \\
  	c_{2} & 0 & a_{22} & 0          \\
  	c_{3} & 0 & a_{32} & a_{33}  \\ \hline
  	         & 0 & a_{32} & a_{33}
  \end{array}
\end{equation}
The problem is then to find the coefficients $\{ \tilde{a}_{21}, \tilde{a}_{31}, \tilde{a}_{32}, a_{22}, a_{32}, a_{33} \}$ satisfying the constraints in Definition~\ref{def:PD-IMEX} while maximizing
\begin{equation}
  c_{\mbox{\tiny Sch}} = \min \Big\{\, \dfrac{c_{20}}{\tilde{c}_{20}},\, \dfrac{c_{30}}{\tilde{c}_{30}},\, \dfrac{c_{32}}{\tilde{c}_{32}} \,\Big\}.  
\end{equation}
%\begin{align*}
%\text{maximize} ~ c_{sch} = &\min \{ \dfrac{c_{20}}{\tilde{c}_{20}}, \dfrac{c_{30}}{\tilde{c}_{30}}, \dfrac{c_{32}}{\tilde{c}_{32}}\}\\
% \text{under these constraints:} \qquad & \\
% \text{(GSA)} \qquad\ &\tilde{w}_{i} = \tilde{a}_{3i}, \quad w_{i} = a_{3i}, ~\text{for} ~ i = 1,2,3\\
% \text{(consistency)} \qquad\ & \sum_{i=1}^{3}  \tilde{w}_{i} = 1, \quad \sum_{i=1}^{3} w_{i} = 1\\ 
% \text{(2nd-order streaming accuracy)} \qquad\ & \sum_{i=1}^{3}  \tilde{w}_{i}\tilde{c}_{i} = 1/2, \\
% \text{(positivity-preserving)} \qquad & a_{ii} > 0, c_{i0} \geq 0, \tilde{c}_{i0} \geq 0, ~\text{for} ~ i = 2,3 \\
% & c_{32} \geq 0, \tilde{c}_{32} \geq 0, \\
% \text{(diffusion accurate)} \qquad & e^{T}_iA^{-1}\tilde{A}e = 1, ~\text{for} ~ i = 2,3, \\
% \text{(affordable)} \qquad & c_{sch} > 0.
%\end{align*}
Imposing the equality constraints (i.e., Eqs.~\eqref{eq:implicitConsistency}, \eqref{eq:orderConditionsEx}, and \eqref{eq:diffusionCondition}), the double Butcher tableau can be written in terms of two parameters ($x$ and $y$) as
\begin{equation}
  \begin{array}{c | c c c}
  	     0       & 0            & 0 & 0 \\
  	\frac{1}{2x} & \frac{1}{2x} & 0 & 0 \\
  	     1       & 1-x          & x & 0 \\ \hline
  	             & 1-x          & x & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0       & 0 & 0            & 0 \\
  	\frac{1}{2x} & 0 & \frac{1}{2x} & 0 \\
  	     1       & 0 & 1-y          & y \\ \hline
  	             & 0 & 1-y          & y
  \end{array}
\end{equation}
Computing the relevant coefficients in Eq.~\eqref{eq:positivityCoefficientsARS}, we find $c_{20}=1$, $c_{30}=1-2x(1-y)$, $c_{32}=2x(1-y)$, $\tilde{c}_{20}=\f{1}{2x}$, $\tilde{c}_{30}=(y-x)$, and $\tilde{c}_{32}=(1-y)$, so that
\begin{equation}
  c_{\mbox{\tiny Sch}} = \min \Big\{\, 2x,\, \f{1-2x(1-y)}{y-x},\, 2(1-y) \,\Big\}.  
\end{equation}
The positivity-preserving property follows from imposing the inequality constraints $a_{22},a_{33}>0$, $c_{20},c_{30},c_{32}\ge0$, and $\tilde{c}_{20},\tilde{c}_{30},\tilde{c}_{32}\ge0$, which imply
\begin{equation}
  0 < x \le y
  \quad\text{and}\quad
  0 < y \le 1.
\end{equation}
We chose $x=\f{1}{2}$, so that the explicit part of the IMEX scheme is equivalent to the optimal second-order SSP-RK scheme in \cite{gottlieb_etal_2001} (SSPRK2 in \ref{app:butcherTables}).  
Then, $y=\f{1}{2} + \epsilon$, where $\epsilon \in [0, \frac{1}{2})$ and $c_{\mbox{\tiny Sch}} = 1 - 2\epsilon$, results in the PD-ARS IMEX scheme.  
Setting $\epsilon = 0$ gives the optimal scheme with $c_{\mbox{\tiny Sch}} = 1$.

\clearpage

\section{Nonexistence of a three-stage PD-IMEX Scheme of Type~A}
\label{app:noTypeA}

Here we prove that a PD-IMEX scheme (i.e., conforming to Definition~\ref{def:PD-IMEX}) of type~A does not exist.  
First, for a three-stage, GSA IMEX scheme of type~A the double Butcher tableau is
\begin{equation*}
  \begin{array}{c | c c c}
  	      0       & 0              & 0              & 0 \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0              & 0 \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0 \\ \hline
  	              & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
    c_{1} & a_{11} & 0      & 0      \\
  	c_{2} & a_{21} & a_{22} & 0      \\
  	c_{3} & a_{31} & a_{32} & a_{33} \\ \hline
  	      & a_{31} & a_{32} & a_{33}
  \end{array}.
\end{equation*}
First we consider the equality constraints.  
Consistency of the implicit coefficients and second-order accuracy in the streaming limit (Eqs.~\eqref{eq:implicitConsistency} and \eqref{eq:orderConditionsEx}, respectively) give
\begin{align}
  a_{31} + a_{32} + a_{33} = 1, \quad \tilde{a}_{31} + \tilde{a}_{32} = 1, \quad \text{and}\quad \tilde{a}_{32}\,\tilde{a}_{21} =  \f{1}{2}.
  \label{eq:1stOrderGeneral}
\end{align}
Accuracy in the diffusion limit (Eq.~\eqref{eq:diffusionCondition}) requires 
\begin{align}
  \tilde{a}_{21} = a_{22}
  \quad\text{and}\quad 
  -\f{a_{32} \tilde{a}_{21}}{a_{22}a_{33}} + \f{1}{a_{33}} =1.
\label{eq:diffusionAccurate}
\end{align}
The first constraint in Eq.~\eqref{eq:1stOrderGeneral} together with the constraints in Eq.~\eqref{eq:diffusionAccurate} gives
\begin{align}
  a_{31} = 0.
  \label{eq:a31is0}
\end{align}
The positivity-preserving property requires $a_{11}, a_{22}, a_{33}>0$ and the inequality constraints
\begin{align*}
  & c_{20} = 1 - \frac{a_{21}}{a_{11}} \geq 0, \quad c_{30} = 1-\frac{a_{31}}{a_{11}} + \frac{a_{32}a_{21}}{a_{22}a_{11}} - \frac{a_{32}}{a_{22}}\geq 0, \\
  & c_{21} = \frac{a_{21}}{a_{11}} \geq 0, \quad c_{31} = \frac{a_{31}}{a_{11}} - \frac{a_{32}a_{21}}{a_{22}a_{11}} \geq 0, \quad c_{32} = \frac{a_{32}}{a_{22}} \geq 0,\\
  & \tilde{c}_{21} = \tilde{a}_{21} \geq 0, \quad \tilde{c}_{31} = \tilde{a}_{31} - \frac{a_{32}\tilde{a}_{21}}{a_{22}} \geq 0, \quad \tilde{c}_{32} = \tilde{a}_{32} \geq 0.
\end{align*}
Since $a_{31} = 0$, we have
\begin{equation*}
  c_{31} = - \frac{a_{32}a_{21}}{a_{22}a_{11}}.  
\end{equation*}
Since $a_{21} \geq 0$, $a_{32} \geq 0$, $a_{22} > 0$ and $a_{11} >0$, we have $c_{31} \leq 0$, and the inequality constraints hold only for $c_{31} = 0$.
Unfortunately, $c_{31} = 0$ gives $c_{\mbox{\tiny Sch}} = \min \Big\{\,\f{c_{21}}{\tilde{c}_{21}}, \f{c_{31}}{\tilde{c}_{31}}, \f{c_{32}}{\tilde{c}_{32}}\Big\} = 0$.
Therefore, a three-stage PD-IMEX scheme (Definition~\ref{def:PD-IMEX}) of type~A does not exist.

>>>>

\section{Construction of IMEX Scheme PD-ARS}
\label{app:PD-ARS}

Here we construct a three-stage PD-IMEX scheme of Type~ARS, conforming to Definition~\ref{def:PD-IMEX}.  
We refer to the resulting IMEX scheme as PD-ARS.  
The double Butcher tableau is
\begin{equation}
  \begin{array}{c | c c c}
  	         0           & 0                 & 0                   & 0  \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0                   & 0  \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0  \\ \hline
  	                   & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0  & 0  & 0         & 0          \\
  	c_{2} & 0 & a_{22} & 0          \\
  	c_{3} & 0 & a_{32} & a_{33}  \\ \hline
  	         & 0 & a_{32} & a_{33}
  \end{array}
\end{equation}
The problem is then to find the coefficients $\{ \tilde{a}_{21}, \tilde{a}_{31}, \tilde{a}_{32}, a_{22}, a_{32}, a_{33} \}$ satisfying the constraints in Definition~\ref{def:PD-IMEX} while maximizing
\begin{equation}
  c_{\mbox{\tiny Sch}} = \min \Big\{\, \dfrac{c_{20}}{\tilde{c}_{20}},\, \dfrac{c_{30}}{\tilde{c}_{30}},\, \dfrac{c_{32}}{\tilde{c}_{32}} \,\Big\}.  
\end{equation}
Imposing the equality constraints (i.e., Eqs.~\eqref{eq:implicitConsistency}, \eqref{eq:orderConditionsEx}, and \eqref{eq:diffusionCondition}), the double Butcher tableau can be written in terms of two parameters ($x$ and $y$) as
\begin{equation}
  \begin{array}{c | c c c}
  	     0       & 0            & 0 & 0 \\
  	\frac{1}{2x} & \frac{1}{2x} & 0 & 0 \\
  	     1       & 1-x          & x & 0 \\ \hline
  	             & 1-x          & x & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
  	     0       & 0 & 0            & 0 \\
  	\frac{1}{2x} & 0 & \frac{1}{2x} & 0 \\
  	     1       & 0 & 1-y          & y \\ \hline
  	             & 0 & 1-y          & y
  \end{array}
\end{equation}
Computing the relevant coefficients in Eq.~\eqref{eq:positivityCoefficientsARS}, we find $c_{20}=1$, $c_{30}=1-2x(1-y)$, $c_{32}=2x(1-y)$, $\tilde{c}_{20}=\f{1}{2x}$, $\tilde{c}_{30}=(y-x)$, and $\tilde{c}_{32}=x$, so that
\begin{equation}
  c_{\mbox{\tiny Sch}} = \min \Big\{\, 2x,\, \f{1-2x(1-y)}{y-x},\, 2(1-y) \,\Big\}.  
\end{equation}
The positivity-preserving property follows from imposing the inequality constraints $a_{22},a_{33}>0$, $c_{20},c_{30},c_{32}\ge0$, and $\tilde{c}_{20},\tilde{c}_{30},\tilde{c}_{32}\ge0$, which imply
\begin{equation}
  0 < x \le y
  \quad\text{and}\quad
  0 < y \le 1.
\end{equation}
We chose $x=\f{1}{2}$, so that the explicit part of the IMEX scheme is equivalent to the optimal second-order SSP-RK scheme in \cite{gottlieb_etal_2001} (SSPRK2 in \ref{app:butcherTables}).  
Then, $y=\f{1}{2} + \epsilon$, where $\epsilon \in [0, \frac{1}{2})$ and $c_{\mbox{\tiny Sch}} = 1 - 2\epsilon$, results in the PD-ARS IMEX scheme.  
Setting $\epsilon = 0$ gives the optimal scheme with $c_{\mbox{\tiny Sch}} = 1$.

\section{Nonexistence of Three-Stage PD-IMEX Scheme of Type~A}
\label{app:noTypeA}

Here we prove that a PD-IMEX scheme (i.e., conforming to Definition~\ref{def:PD-IMEX}) of type~A does not exist.  
First, for a three-stage, GSA IMEX scheme of type~A the double Butcher tableau is
\begin{equation*}
  \begin{array}{c | c c c}
  	      0       & 0              & 0              & 0 \\
  	\tilde{c}_{2} & \tilde{a}_{21} & 0              & 0 \\
  	\tilde{c}_{3} & \tilde{a}_{31} & \tilde{a}_{32} & 0 \\ \hline
  	              & \tilde{a}_{31} & \tilde{a}_{32} & 0
  \end{array}
  \qquad
  \begin{array}{c | c c c}
    c_{1} & a_{11} & 0      & 0      \\
  	c_{2} & a_{21} & a_{22} & 0      \\
  	c_{3} & a_{31} & a_{32} & a_{33} \\ \hline
  	      & a_{31} & a_{32} & a_{33}
  \end{array}.
\end{equation*}
First we consider the equality constraints.  
Consistency of the implicit coefficients and second-order accuracy in the streaming limit (Eqs.~\eqref{eq:implicitConsistency} and \eqref{eq:orderConditionsEx}, respectively) give
\begin{align}
  a_{31} + a_{32} + a_{33} = 1, \quad \tilde{a}_{31} + \tilde{a}_{32} = 1, \quad \text{and}\quad \tilde{a}_{32}\,\tilde{a}_{21} =  \f{1}{2}.
  \label{eq:1stOrderGeneral}
\end{align}
Accuracy in the diffusion limit (Eq.~\eqref{eq:diffusionCondition}) requires 
\begin{align}
  \tilde{a}_{21} = a_{22}
  \quad\text{and}\quad 
  -\f{a_{32}\,\tilde{a}_{21}}{a_{22}\,a_{33}} + \f{1}{a_{33}} =1.
\label{eq:diffusionAccurate}
\end{align}
The first constraint in Eq.~\eqref{eq:1stOrderGeneral} together with the constraints in Eq.~\eqref{eq:diffusionAccurate} gives
\begin{align}
  a_{31} = 0.
  \label{eq:a31is0}
\end{align}
The positivity-preserving property requires $a_{11}, a_{22}, a_{33}>0$ and the inequality constraints
\begin{align*}
  & c_{20} = 1 - \frac{a_{21}}{a_{11}} \geq 0, \quad c_{30} = 1-\frac{a_{31}}{a_{11}} + \frac{a_{32}\,a_{21}}{a_{22}\,a_{11}} - \frac{a_{32}}{a_{22}}\geq 0, \\
  & c_{21} = \frac{a_{21}}{a_{11}} \geq 0, \quad c_{31} = \frac{a_{31}}{a_{11}} - \frac{a_{32}\,a_{21}}{a_{22}\,a_{11}} \geq 0, \quad c_{32} = \frac{a_{32}}{a_{22}} \geq 0,\\
  & \tilde{c}_{21} = \tilde{a}_{21} \geq 0, \quad \tilde{c}_{31} = \tilde{a}_{31} - \frac{a_{32}\,\tilde{a}_{21}}{a_{22}} \geq 0, \quad \tilde{c}_{32} = \tilde{a}_{32} \geq 0.
\end{align*}
Since $a_{31} = 0$, we have
\begin{equation*}
  c_{31} = - \frac{a_{32}a_{21}}{a_{22}a_{11}}.  
\end{equation*}
Since $a_{21} \geq 0$, $a_{32} \geq 0$, $a_{22} > 0$ and $a_{11} >0$, we have $c_{31} \leq 0$, and the inequality constraints hold only for $c_{31} = 0$.
Unfortunately, $c_{31} = 0$ gives $c_{\mbox{\tiny Sch}} = \min \Big\{\,\f{c_{21}}{\tilde{c}_{21}}, \f{c_{31}}{\tilde{c}_{31}}, \f{c_{32}}{\tilde{c}_{32}}\Big\} = 0$.
Therefore, a three-stage PD-IMEX scheme (Definition~\ref{def:PD-IMEX}) of type~A does not exist.
